The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-10-08 article.tex
Next iteration

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\gio{According to my fine graining approach we should at each step deepen every small section. At a certain level I think we can start to re-coarse-grain what we wrote to create the abstract.}
\richel{I enjoy this approach! Did some minor fine-graining}
\richel{Have you already looked up for a target journal? I know how a journal's constraints have helped me 
\gio{Honestly I have literally no idea how to select a good journal for this kind of article.}
in writing an article, for example, by having a maximum number of pictures}

@@

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model in order to perform a fair comparison. 
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations. %option 1
%i.e. the same number of taxa at the present. %option 2
To obtain these twin trees, 
\richel{
  I suggest to first start with the equal number of taxa,
  and the calculation of the speciation rates first.
}
\gio{@richel: both assumptions seem to make sense to me. If possible I would run both and see which one performs better. For now let's stick to option 1 in the manuscript.}
we impose
that the expected number of mutations in an MBD tree, $m_{MBD}$ equals
the expected number of mutations in a BD tree, $m_{BD}$:

@@

\gio{I think this should definitely go to the methods}
\richel{I put it there for use, hopefully at a spot you liked :-)}

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{I don't agree. It is expected if you use an expected number of species over time. If you have the correct $n(t)$ in principle you should be able to get the exact number of mutations.}
with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  above stood 'crown age -T'. 
  I feel that a crown age is a positive number,
  but I know you have had a good reason.
  Perhaps better would be to write something explicit like:
  t\_now - t\_crown = T.
  Looking forward for a better suggestion than mine :-)
}
\gio{We can specify that the crown age is $-T$ where $T$ is a positive number (which should be already clear, I guess, if we report it as $-T$). In my opinion this makes everything nicer when we have to write down the integral which should, in principle, be $\int_{-T}^{0} f(t-(-T))$ but using integral properties can be rewritten as $\int_{0}^{T} f(t)$. Having a definite integral starting from $0$ always looks better ^^.}

@@

%It can be easily seen considering that $\rho$ is the number of mutations
%per unit time per site. For this reason it's needed to multiply by
%time and number of sites. 
\gio{This feels kind of a repetitions of what we wrote before the formula. I comment it. We can think of reinsert it afterwards, if needed (see comment above).}
\richel{
  I suggest to remove such commented-out lines. 
  Although I sometimes get attached to my sentences,
  they clog up the document by non-info and 
  I usually delete them anyways in the end.
  I cannot remember ever regretting this (would I, 
  I could find it in the git history).
  Would you agree?
}
\gio{Agree. Remove them. CLOSED}

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}
\richel{
  I see you use angle bracket as a notation for the expected
  value. I usually see 'E(x)' as the expected value for 'x',
  and this is used at the beloved https://en.wikipedia.org/wiki/Expected\_value.
  What are the reasons you prefer the notation with the angle brackets?  
}
\gio{Because the capital E is inelegant, plebeian and boor. And we want to be classy, don't we? :D}

@@

\gio{My doubt is if we need to use $m_{MBD}$ for the single tree or the same
quantity averaged on the full MBD dataset $<m_{MBD}>$.
Do you think is better to use the
individual $m_{MBD}$ for each tree or the average across the whole
dataset?}
\richel{
  I think a per-tree calculation of the mutation rates 
  is the best we can do. As there was some noise between us above,
  I think the iteration after the next will allow us to get a better
  idea about this
}
\gio{I am still very debated about this. I actually would like to ask also Rampal's opinion.
}
\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{mbd.jpg}
  \caption{
    How to create twin trees and alignments. From a focal MBD tree, a twin tree is produced as 
    such: (1) estimate the $\lambda_{BD}$ to get the same expected number of tips, (2) simulate a BD tree with that amount of tips (discard trees with different number of tips), (3) estimate a mutation rate to get an alignment with the same expected number of mutations, (4) simulate alignments with that amount of mutations (discard those that don't, the picture shows an alignment that should be discarded) 
  }
\end{figure}

@@

\item So far we have simulated two datasets of trees under the two models: $\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
2) Writing down things to tidy up manuscript and thoughts I just realized that actually the "model" is the same for the two posteriors. In fact with the word "model" Maturana et al. refer to the prior used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually use the Bayes factor for model selection?}
%Now we have two datasets  of posteriors to compare, one for the BD model and one for the MBD model.

@@

\gio{I am running out of letters :(}
\richel{
  I would love to describe this more concrete. For example, when do
  we say something has an effect? If we avoid making such judgements,
  how will we visualize?
}

@@

\richel{I removed the Bayes factor text. It is useful when letting BEAST2
pick more/overly complex models and see if that more complex model fits the
data better (penalized by its increased complexity, similar to the AIC). It has
its uses, but I am unsure if we already want to discuss this now
or first focus on the proper tree twinning}
\gio{But after we describe the tree twinning we will have to describe the method we use for model selection anyways. BF and nLTT are mutually exclusive choices, right?}
\gio{The point is: what method do we choose? According to the choice we have to describe that. I'd also like to know Rampal's opinion on that. In principle I like more the BF method cause it looks more solid while the nLTT looks a bit like the last resort to me.}

@@

\item
\richel{
  I guess you know I am a fan of the Open Science Framework,
  in which you first register you work before you do the experiment
  (note: I will do some small pilots to estimate the complete time
  of the experiment). I think it is the proper and superior science,
  which helps us against writing down bullshit stories after having
  obtained the results (e.g. 'We expected A and indeed found it!').
  It also helps me structure my work: first think
  deeply about the experiment, then do it (instead of the mixing
  up the two phases). What are your thoughts on that?
}
\gio{I agree on the principle but I don't fully agree on the open science procedure. Sometimes you do find something good while you're looking for something else. That happened many times in the history of science. This doesn't necessarily implies that we will end up p-hacking the results. I think we need to be open to new findings that might be not forecasted by our hypothesis. I think this is the good way to do science: be critical with our results but not imposing that reality will follow the theory.}
\item

>>>>

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\richel{Have you already looked up for a target journal?} 
\gio{Honestly I have literally no idea how to select a good journal for this kind of article.}
\richel{I suggest you ask Rampal}

@@

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model in order to perform a fair comparison. 
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations and 
the same number of taxa at the present.
To obtain these twin trees, 
\richel{
  I suggest to first start describing how to create the twin tree,
  calculation of the speciation rates, then the twin alignment.
}
\gio{@richel: both assumptions seem to make sense to me. 
  If possible I would run both and see which one performs better. For now
  let's stick to option 1 in the manuscript.
}
\richel{
  Sorry for the confusion: the previous comment was about the order
  of describing out methods in this manuscript. As of the methods,  
  I really meant to create twin trees with 
  both equal number of taxa and equal expected number of mutations.
}

we impose
that the expected number of mutations in an MBD tree, $m_{MBD}$ equals
the expected number of mutations in a BD tree, $m_{BD}$:

@@

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{
  I don't agree. It is expected if you use an expected number 
  of species over time. If you have the correct $n(t)$ in principle you 
  should be able to get the exact number of mutations.
}
\richel{
  If you have the correct $n(t)$ you can indeed calculate exactly what
  the number of mutations will be, if and only if the mutation rate is
  zero. Except for this travial case, I am curious about how you
  calculate this. What is the precise number of mutations on a DNA
  sequence of 1000 base pairs over one time unit with an 
  expected average per-nucleotide
  mutation rate of once per time? I would say one expects on average 1000 mutations
  to occur. What would your claim be?
}

with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  above stood 'crown age -T'. 
  I feel that a crown age is a positive number,
  but I know you have had a good reason.
  Perhaps better would be to write something explicit like:
  t\_now - t\_crown = T.
  Looking forward for a better suggestion than mine :-)
}
\gio{
  We can specify that the crown age is $-T$ where $T$ is a positive number 
  (which should be already clear, I guess, 
  if we report it as $-T$). In my opinion this makes everything 
  nicer when we have to write down the integral which should, 
  in principle, be $\int_{-T}^{0} f(t-(-T))$ but using integral properties 
  can be rewritten as $\int_{0}^{T} f(t)$. Having a definite integral 
  starting from $0$ always looks better ^^.
}
\richel{
  I see, in a conversion you would say 'the crown age of the phylogeny is 
  minus ten million years', where I would the say 'the crown age of the 
  phylogeny is ten million years' to say exactly the same thing!
  To try to resolve this, I checked one (actually two) of Rampal's papers (
    'Diversity-dependence brings molecular phylogenies closer to agreement with the fossil record',
    by Rampal, Bart Haegeman, Tanja Stadler, Tracy Aze, Paul N. Pearson, Andy Purvis, Albert B. Phillimore
  ). Rampal avoids this point altogether: '... crown age $t_c = t_p − t_1$'
  and '... and crown ages (5, 10 and 15 Myr)'. I suggest we do the same in
  both writing and derivations: he is even more famous than us :-) 
}

@@

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}
\richel{
  I see you use angle bracket as a notation for the expected
  value. I usually see 'E(x)' as the expected value for 'x',
  and this is used at the beloved https://en.wikipedia.org/wiki/Expected\_value.
  What are the reasons you prefer the notation with the angle brackets?  
}
\gio{
  Because the capital E is inelegant, plebeian and boor. 
  And we want to be classy, don't we? :D
}
\richel{
  If the angle brackets are the new hip kids in mathematics town, sure,
  we'll use those!
}

@@

\gio{My doubt is if we need to use $m_{MBD}$ for the single tree or the same
quantity averaged on the full MBD dataset $<m_{MBD}>$.
Do you think is better to use the
individual $m_{MBD}$ for each tree or the average across the whole
dataset?}
\richel{
  I think a per-tree calculation of the mutation rates 
  is the best we can do. As there was some noise between us above,
  I think the iteration after the next will allow us to get a better
  idea about this
}
\gio{
  I am still very debated about this. I actually would like to ask also Rampal's opinion.
}
\richel{
  Sounds like a good idea!
}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{mbd.jpg}
  \caption{
    How to create twin trees and alignments. From a focal MBD tree, a twin tree is produced as 
    such: (1) estimate the $\lambda_{BD}$ to get the same expected number of tips, (2) simulate a BD tree with that amount of tips (discard trees with different number of tips), (3) estimate a mutation rate to get an alignment with the same expected number of mutations, (4) simulate alignments with that amount of mutations (discard those that don't, the picture shows an alignment that should be discarded) 
  }
\end{figure}

@@

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
}
\richel{
  I definitely get your point! 
  I would prefer something expressive over a number, perhaps a fancy letter, like
  for example $\mathbb{B}$ and $\mathbb{M}$ for a Birth-Death and Multiple-Birth Death. 
}
\gio{
2) Writing down things to tidy up manuscript and thoughts I just realized that 
  actually the "model" is the same for the two posteriors. 
  In fact with the word "model" Maturana et al. refer to the prior 
  used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually 
  use the Bayes factor for model selection?
}
\richel{
  Correct! The assumed model of the posterior is a Birth-Death model. To do
  model selection, we could assume a different nucleotide substitution 
  model (GTR instead of JC69) and/or a different clock model (relaxed clock,
  instead of a strict clock). I hope this clears things up. I will add
  'generative model' for the model (BD or MBD) that created a phylogeny
  and 'inferential model' for the model used in generating a posterior.
}

@@

\gio{I am running out of letters :(}
\richel{Haha! I suggest to use the same equation and symbols 
  as equation 1 in
  the nLTT article of Janzen, Hoehna and Etienne, 2015:
}

$$
\delta nLTT = \int_{0}^{1} \abs{nLTT_1(t) - nLTT_2{t}} dt$
$$

@@

\richel{
  I removed the Bayes factor text. It is useful when letting BEAST2
  pick more/overly complex models and see if that more complex model fits the
  data better (penalized by its increased complexity, similar to the AIC). It has
  its uses, but I am unsure if we already want to discuss this now
  or first focus on the proper tree twinning
}
\gio{
  But after we describe the tree twinning we will have to describe 
  the method we use for model selection anyways. 
  BF and nLTT are mutually exclusive choices, right?
}
\richel{
  BF and nLTT are not mutually exclusive choices, but do complementary things:
  nLTT measures the difference between the 'true' tree and the inferred trees,
  which is the most important thing: we are interested in the error BEAST2
  makes when using an overly simple (single-)BD tree prior.
  With the BF we can compare BEAST2 posteriors that use (the same
  tree prior, but) different site models (e.g. JC69, the simplest, 
  versus GTR, the most complex). Although we know we generated the DNA
  alignment with JC69, there are some hints that the GTR model is the
  superior choice to do inference with.
}
\gio{
  The point is: what method do we choose? 
  According to the choice we have to describe that. 
  I'd also like to know Rampal's opinion on that. 
  In principle I like more the BF method cause it looks 
  more solid while the nLTT looks a bit like the last resort to me.
}
\richel{
  I hope this is now more clear. If not, I'll happily
  elaborate :-)
}

@@

\item
\richel{
  I guess you know I am a fan of the Open Science Framework,
  in which you first register you work before you do the experiment
  (note: I will do some small pilots to estimate the complete time
  of the experiment). I think it is the proper and superior science,
  which helps us against writing down bullshit stories after having
  obtained the results (e.g. 'We expected A and indeed found it!').
  It also helps me structure my work: first think
  deeply about the experiment, then do it (instead of the mixing
  up the two phases). What are your thoughts on that?
}
\gio{I agree on the principle but I don't fully agree on the open science procedure. Sometimes you do find something good while you're looking for something else. That happened many times in the history of science. This doesn't necessarily implies that we will end up p-hacking the results. I think we need to be open to new findings that might be not forecasted by our hypothesis. I think this is the good way to do science: be critical with our results but not imposing that reality will follow the theory.}
\richel{
  Looks like we are on the same page here: we both dislike making up bullshit 
  stories after the results are known. Great!
}
\item
